Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-12-21 main.tex
add TODOs after review
some more explanations in static analysis
begin extraction of the scenario
add sets of actions
fixes and suggestions of Morgan

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

@@

% Pretty references
\usepackage{prettyref}
\newrefformat{def}{Def.~\ref{#1}}
\newrefformat{fig}{Fig.~\ref{#1}}
\newrefformat{lem}{Lemma~\ref{#1}}
\newrefformat{th}{Theorem~\ref{#1}}
\newrefformat{co}{Corollary~\ref{#1}}
\newrefformat{sec}{Sect.~\ref{#1}}
\newrefformat{ssec}{Subsect.~\ref{#1}}
\newrefformat{suppl}{Appendix~\ref{#1}}
\newrefformat{cr}{Condition~\ref{#1}}
\newrefformat{eq}{Eq.~\eqref{#1}}
\def\pref{\prettyref}

@@

@@

\begin{abstract}
The Process Hitting is a recently introduced framework designed for the modelling of concurrent systems.
Its originality lies in a compact representation of both components of the model and its corresponding actions:
each action can modify the status of a component, and is conditioned by the status of at most one other component.
This allowed to define very efficient static analysis based on local causality to compute reachability properties.
However, in the case of cooperations between components (for example, when two components are supposed to interact with a third one only when they are in a given configuration), the approach leads to an over-approximated interleaving between actions, because of the pure asynchronous semantics of the model.

To address this issue, we propose an extended definition of the framework, including priority classes for actions.
In this paper, we focus on a restriction of the Process Hitting with two classes of priorities and a specific behaviour of the components, that is sufficient to tackle the aforementioned problem of cooperations.
We develop a new refinement for the under-approximation of the static analysis to give accurate results for this class of Process Hitting models.
Then we show that this class of models is sufficient to represent any
Process Hitting model with an arbitrary number of classes of priorities,
and even any Asynchronous Discrete Networks, either Boolean or multivalued.
Our method thus allows to efficiently under-approximate reachability properties
in Asynchronous Discrete Networks,
as it is illustrated on the model-checking of a signalling network of 94
components, which is unprecedented.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

>>>>

\usepackage[UKenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

@@

% Pretty references
\usepackage{prettyref}
\newrefformat{def}{Def.~\ref{#1}}
\newrefformat{fig}{Fig.~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}
\newrefformat{lem}{Lemma~\ref{#1}}
\newrefformat{th}{Theorem~\ref{#1}}
\newrefformat{co}{Corollary~\ref{#1}}
\newrefformat{sec}{Sect.~\ref{#1}}
\newrefformat{ssec}{Subsect.~\ref{#1}}
\newrefformat{suppl}{Appendix~\ref{#1}}
\newrefformat{cr}{Condition~\ref{#1}}
\newrefformat{eq}{Eq.~\eqref{#1}}
\def\pref{\prettyref}

@@

% Ajouts et progression
\def\modMF#1{\textcolor{teal}{#1}}
\def\modLP#1{\textcolor{magenta}{#1}}
\def\modMM#1{\textcolor{blue}{#1}}
\def\modOR#1{\textcolor{olive}{#1}}
%\def\modMF#1{#1} \def\modLP#1{#1} \def\modMM#1{#1} \def\modOR#1{#1}

@@

\begin{abstract}
The Process Hitting is a recently introduced framework designed for the modelling of concurrent systems.
Its originality lies in a compact representation of both components of the model and its corresponding actions:
each action can modify the status of a component, and is conditioned by the status of at most one other component.
This allowed to define very efficient static analysis based on local causality to compute reachability properties.
However, in some cases, this modelling lacks some preemption features
that would allow to model relative time constraints between interactions.

In this paper, we propose an extension of the Process Hitting framework
by using classes of priorities in order to improve its expressivity.
We also show that these priorities allow to represent accurate Boolean gates in order to model
cooperations between components
(that is, for instance, when two components are supposed to interact with a third one
only when they are in a given configuration)
which was not possible before.

Furthermore, we develop in this paper a new refinement for the under-approximation of the static analysis
to give accurate results on Process Hitting models.
This method applies to a specific sub-class of models,
but this class turns out to be sufficiently expressive to represent any
Process Hitting model with an arbitrary number of classes of priorities,
thus guaranteeing a wide field of application of our method.
Finally, we show that Process Hitting allows to accurately model
any Asynchronous Discrete Networks, either Boolean or multivalued.
Our static analysis method thus allows to efficiently under-approximate reachability properties
in Asynchronous Discrete Networks,
as it is illustrated on the model-checking of a signalling network of 94
components, which is unprecedented.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

### 2014-12-21 parts/def-ph.tex
fixes and suggestions of Morgan

\modMF{%
This section aimed at defining the PH framework in a general fashion,
and focused on a more specific class of PH models
that will be used later in this paper.
We gave a definition of the notion of cooperative sort,
a special sort with no explicit biological existence or meaning,
which is used to model the joint action of several components
on another one.
Using this definition, we defined a specific class of PH models
that assign a specific role to the set of actions
of highest priority:
these actions can only be used to update cooperative sorts to ensure that
they are equivalent to Boolean gates.
The static analysis developed in the next section can only be applied
to PH models belonging to this specific class.
However, it has to be noted that this class is in fact as expressive as
the class of all possible PH models with priorities (see \pref{ssec:flattening}).
}%

>>>>

\modMF{%
This section aimed at defining the PH framework in a general fashion,
and focused on a more specific class of PH models
that will be used later in this paper.
We gave a definition of the notion of cooperative sort,
a special sort with no explicit biological existence or meaning,
which is used to model the joint action of several components
on another one.
Using this definition, we defined a specific class of PH models
that assign a specific role to the set of actions
of highest priority:
these actions can only be used to update cooperative sorts to ensure that
they are equivalent to Boolean gates.
Technically, the static analysis developed in the next section has to be applied
to PH models belonging to this specific class with only two sets of priorities.
However, it has to be noted that this class is in fact as expressive as
the class of all possible PH models with priorities (see \pref{ssec:flattening}),
and that the static analysis developed is thus not limited to this sub-class of models.
}%

### 2014-12-21 parts/example.tex
fixes and suggestions of Morgan

\modMF{
This section aims at giving application examples of the static analysis method
that we developed in \pref{sec:sa}.
In \pref{ssec:ex-metazoan}, we apply our method to a small model of the
metazoan segmentation process,
and demonstrate how the flattening of \pref{ssec:flattening}
be used in such a case.
In \pref{ssec:ex-tcrsig}, we apply our method to a large-scale model
of the T-cell receptor signalling pathway
in order to show the scalability of our method.
}

>>>>

\modMF{
This section aims at giving application examples of the static analysis method
that we developed in \pref{sec:sa}.
In \pref{ssec:ex-metazoan}, we apply our method to a small model of the
metazoan segmentation process,
and demonstrate how the flattening of \pref{ssec:flattening}
can be used in such a case.
In \pref{ssec:ex-tcrsig}, we apply our method to a large-scale model
of the T-cell receptor signalling pathway
in order to show the scalability of our method.
}

### 2014-12-21 parts/intro.tex
fixes and suggestions of Morgan

Discrete modelling frameworks for biological networks is an active research field where formal methods have proved that they were very powerful.
Such a work started in the seventies.
It was later enriched in many directions and widely used to elucidate many biological questions.
Among these questions, a major one is to understand precisely how biological systems evolve and behave; why and how they change their usual behaviours, etc.
This leads to questions about the reachability (possible or inevitable) of some states.
The ultimate goal is to discover how it could be possible to prevent biological systems from reaching some pathological states.
\todo{Ajouter des références}

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Prevalent formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component and in their multivalued equivalent, Asynchronous Discrete Networks~\cite{deJong02}, components rarely have more than four levels.

@@

\subsection*{Contributions}
Since our formalism (the Process Hitting) can be considered as a subset of Calculus of Communicating Systems, %is inspired from the $\pi$-calculus,
our work is related to such semantic ramifications of extending traditional process algebras with the concept of priority that allows for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refers to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a static priority structure.
The actions modelling biological processes are divided in several classes of priorities which stand for the different reaction rates or relative preemptions.
Furthermore, the highest class of priority is used to model particular nonbiological actions, designated as “prioritised”,
that are used to encode Boolean functions inside the model.

@@

Whereas the analysis presented in \pref{sec:sa} considers a particular class of
Process Hitting with two classes of priority,
we provide in \pref{sec:encodings} formal encodings of Asynchronous Discrete Networks and general
Process Hitting models with priorities.
Those encodings result in a Process Hitting fulfilling the conditions for our
static analysis and support the broad applicability of our method.

@@

\modMF{
The main addition in this paper compared to \cite{FPMR13-CS2Bio} is
%two theorems, the first
\pref{thm:ordered-ua} that gives a method to cope with successive reachability,
and obtain more conclusive results in the case of several successive objectives.
Two other additions are 
\pref{th:bisimPHP} which states that any Process Hitting model with any number of classes of priorities
can be represented with only two classes of priorities,
thus extending the scope of our method to any possible Process Hitting model,
and \todo{def extraction}
which allows to extract a concrete scenario when our static analysis method
is conclusive.
%Both aim at extending the application range of the static analysis we developed.
}

>>>>

Discrete modelling frameworks for biological networks is an active research field where formal methods have proven that they were very powerful.
Such a work started in the seventies.
It was later enriched in many directions and widely used to elucidate many biological questions.
Among these questions, a major one is to understand precisely how biological systems evolve and behave; why and how they change their usual behaviours, etc.
This leads to questions about the (possible or inevitable) reachability of some states.
The ultimate goal is to discover how it could be possible to prevent biological systems from reaching some pathological states.
\todo{Ajouter des références}

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Prevalent formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction of these frameworks where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component and in their multivalued equivalent, Asynchronous Discrete Networks~\cite{deJong02}, components rarely have more than four levels.

@@

\subsection*{Contributions}
\todo{Compléter les contributions}
Since our formalism (the Process Hitting) can be considered as a subset of Calculus of Communicating Systems, %is inspired from the $\pi$-calculus,
our work is related to such semantic ramifications of extending traditional process algebras with the concept of priority that allows for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refers to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a static priority structure.
The actions modelling biological processes are divided in several classes of priorities which stand for the different reaction rates or relative preemptions.
Furthermore, the highest class of priority is used to model particular nonbiological actions, designated as “prioritised”,
that are used to encode Boolean functions inside the model.

@@

Whereas the analysis presented in \pref{sec:sa} considers a particular class of
Process Hitting with two classes of priority,
we provide in \pref{sec:encodings} a “flattening” of any Process Hitting model into
this class of models,
so that its application is not restricted to it.
We also give in the same section a formal encodings of Asynchronous Discrete Networks,
allowing to efficiently study this kind of widespread models.

@@

\modMF{
The main addition in this paper compared to \cite{FPMR13-CS2Bio} are the following:
\pref{th:bisimPHP} states that any Process Hitting model with any number of classes of priorities
can be represented with only two classes of priorities,
thus extending the scope of our method to any possible Process Hitting model;
\pref{thm:ordered-ua} gives a method to cope with successive reachability,
and obtain more conclusive results in the case of several successive objectives;
\todo{def extraction} allows to extract a concrete scenario when our static analysis method
is conclusive.
%All aim at extending the application range of the static analysis we developed.
}
